Treebeard : An Optimizing Compiler for Tree Based ML Inference

### 2022-04-17 docs/MICRO 2022/probtiling.tex
unif tililng

\subsection{Probability Based Tiling}
\label{sec:ProbTiling}
\subsubsection{Motivation}
Uniform tiling optimizes tree walks assuming that all root to leaf paths in the tree are equally likely. 
However, we find that this is not always the case. We performed inference on the complete training set of each of our benchmarks and counted how many
times each leaf was reached. While for some benchmarks all leaves are roughly equally likely, for others, a small number of leaves are extremely likely 
while the rest aren't reached often. We call trees with a small number of extremely likely leaves \textbf{\emph{leaf biased}}.
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/airline-ohe.stats.train.txt.png}
    \caption{Statistical profile for airline-ohe}
    \label{Fig:AirlineOHEStats}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/epsilon.stats.train.txt.png}
    \caption{Statistical profile for epsilon}
    \label{Fig:EpsilonStats}
\end{figure}

>>>>

\subsection{Probability Based Tiling}
\subsubsection{More Notation}
In order to formulate the probability based tiling algorithm as an optimization problem, we define the following.
\begin{enumerate}
    \item For every leaf $l \in L$, we define $p_l$ as the probability that the leaf $l$ is reached.
    \item For each node $n \in V$, we define the absolute probability $p_v$ as
    \begin{equation}
        p_v = \begin{cases}
        p_l &\text{if $l \in L$}\\
        p_{left(v)} + p_{right(v)} &\text{otherwise}
        \end{cases}
    \end{equation}
    \item For any tree $T$, $\mathcal{C}(T)$ represents the set of all valid tilings of $T$.
    \item For every $v \in V$, we define $S_v$ as the subtree rooted at $v$.
    \item For every $v \in V$, we define $L_v$ as the set of leaves of $S_v$.
    \item For a every tile $T_i$, we define $root(T_i)$ as the node $v \in T_i$ such that $v$ has no incoming edges from any other node $u \in T_i$.
    \item For a tile $T_i$, $out(T_i) \subseteq E$ is the set of edges $(u, v)$ such that $u \in T_i$ and $v \notin T_i$.
\end{enumerate}

\label{sec:ProbTiling}
\subsubsection{Motivation}
Uniform tiling optimizes tree walks assuming that all root to leaf paths in the tree are equally likely. 
However, we find that this is not always the case. We performed inference on the complete training set of each of our benchmarks and counted how many
times each leaf was reached. While for some benchmarks all leaves are roughly equally likely, for others, a small number of leaves are extremely likely 
while the rest aren't reached often. We call trees with a small number of extremely likely leaves \textbf{\emph{leaf biased}}.
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/airline-ohe.stats.train.txt.png}
    \caption{Statistical profile for airline-ohe}
    \label{Fig:AirlineOHEStats}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/epsilon.stats.train.txt.png}
    \caption{Statistical profile for epsilon}
    \label{Fig:EpsilonStats}
\end{figure}

### 2022-04-17 docs/MICRO 2022/tiling.tex
unif tililng

\subsection{Notation}
\TODO{Notation needs to be introduced in the background section}
We represent a decision tree by $T = (V, E, r)$ where $V$ is the set of nodes, $E$ the set of edges and
$r \in V$ is the root. For each node $n \in V$, we define the following.
\begin{enumerate}
    \item $threshold(n) \in \mathbb{R}$, the threshold value for $n$.
    \item $featureIndex(n) \in \mathbb{N}$, the feature index for $n$.
    \item $left(n) \in V$, the left child of $n$ or $\emptyset$ if $n$ is a leaf. If $left(n) \neq \emptyset$, then $(n, left(n)) \in E$.
    \item $right(n) \in V$, the right child of $n$ or $\emptyset$ if $n$ is a leaf. If $right(n) \neq \emptyset$, then $(n, right(n)) \in E$.
\end{enumerate}
We use $L \subseteq V$ to denote the set of leaves. % subseteq because tree could have single node

\subsection{Tiling}
\label{sec:Tiling}
% Treebeard vectorizes tree walks by grouping nodes of a decision tree into \textbf{\emph{tiles}}. The nodes in a tile are evaluated concurrently using vector instructions. Once the nodes of the current tile are evaluated, a look up table is used to compute which child of the current tile to move to next.
Treebeard groups nodes of the decision tree into \textbf{\emph{tiles}}. Tiling provides two benefits. 
\begin{enumerate}
  \item It allows the compiler to generate vector code to traverse trees. Section \ref{sec:Vectorization} describes how Treebeard does this.
  \item It enables spatial locality improvements by grouping together nodes that are likely to be accessed together. 
\end{enumerate}
Once nodes are grouped into tiles, an $n$-ary tree whose nodes are tiles is constructed. Treebeard then generates 
optimized code to walk this tree. The listing below shows at a high level how a tiled tree is walked (This is not 
true IR, but presented for clarity). 
\begin{lstlisting}[style=c++]
  ResultType Prediction_Function(...) {
    // ...
    Tile t = getRootTile(tree)
    while (!isLeaf(tree, t)) do {
      // Evaluate conditions of all nodes in the tile
      predicates = evaluateTilePredicates(t, rows[i])

      // Move to the correct child of the current tile
      t = getChildTile(tree, t, predicates) 
    }
    treePrediction = getLeafValue(t)
    // ...
  }  
\end{lstlisting}

To compute the prediction of the tree, the predicates of all nodes in the tile are computed simultaneously (line 6). 
Then, the computed predicate values are used to determine which child of the current tile to move to (in the 
$n$-ary tree). This section presents the details of tiling and how Treebeard's general tiling infrastructure 
can be used to develop tiling algorithms with different objectives (sections \ref{sec:UnifTiling}) and \ref{sec:ProbTiling}).
The details of how Treebeard lowers predicate evaluation and moving to the correct child to use vector instructions 
are described in section \ref{sec:Vectorization}.

\subsection{Tiles and Tree Tiling}
\label{sec:ValidTiling}
A \textbf{\emph{tile}} is a collection of connected non-leaf nodes of a decision tree. The path connecting any pair of nodes
in the tile must be contained fully within the tile. The \textbf{\emph{tile size}} $n_t$ is the number of nodes contained in a tile.

A \textbf{\emph{tiling}} $\mathcal{T}$ of the tree $T = (V, E, r)$ with tile size $n_t$ is a partition $\{ T_1, T_2, ... ,T_m \}$ of the set $V$ such that 
\begin{enumerate}
    \item $T_1 \cup T_2 ... \cup T_m = V$
    \item $T_i \cap T_j = \emptyset$ for all $i, j \in [1, m]$ and $i \neq j$
    \item Tiles are \textbf{connected}, i.e. for an $u, v \in T_i$, there is a (undirected) path connecting $u$ and $v$ fully contained in $T_i$.
\end{enumerate}

We say a tiling $\mathcal{T}$ is a \textbf{\emph{valid tiling}} iff
\begin{enumerate}
  \item $|T_i| \leq n_t$ for all $i \in [1, m]$
  \item $\forall l \in L$ : $l \in T_i \rightarrow v \notin T_i \;\; \forall v \in V \backslash \{l\}$
  \item Tiles are \textbf{maximal}, i.e. if $|T_i| < n_t$, then there is no $v \in V\backslash \{ T_i \cup L \}$ such that $(u, v) \in E$ for some $u \in T_i$. 
\end{enumerate}

\subsection{Tiled Trees}
A tiling transformation communicates the tiling to the Treebeard infrastructure by assigning a tile ID to each node in the decision tree. Using these tile IDs, Treebeard checks the validity of the tiling and then contructs a tree whose nodes are tiles. We call this tree the \textbf{\emph{tree of tiles}}. \TODO{We need a better name for this}
Figure \ref{Fig:ValidTilingTileSize3} shows a valid tiling with tile size 3 and the tree of tiles constructed by Treebeard. Three nodes are grouped into each of the tiles $t_1$ and $t_2$ as shown. Each tile is collapsed into a single node in the tree of tiles. However, each leaf in the original tree becomes a leaf in the tree of tiles.

@@

Treebeard maintains the following invariants.
\begin{enumerate}
  \item All tiles in a tree are the same size $n_t$. If the tiling produces any smaller tiles, these are padded by inserting dummy nodes to make them the required size.
  \item Nodes within tiles are always ordered in level order and left to right within a level. The numbering of the nodes in the above diagram shows this node order.
  \item Children of a node are numbered from left to right (regardless of level). For example, $l_1$ is the first child of $t_1$, $l_2$ is the second and so on.
\end{enumerate}

>>>>

\subsection{Notation}
\TODO{Notation needs to be introduced in the background section, Tiling should be the first subsection}
We represent a decision tree by $T = (V, E, r)$ where $V$ is the set of nodes, $E$ the set of edges and
$r \in V$ is the root. For each node $n \in V$, we define the following.
\begin{enumerate}
    \item $threshold(n) \in \mathbb{R}$, the threshold value for $n$.
    \item $featureIndex(n) \in \mathbb{N}$, the feature index for $n$.
    \item $left(n) \in V$, the left child of $n$ or $\emptyset$ if $n$ is a leaf. If $left(n) \neq \emptyset$, then $(n, left(n)) \in E$.
    \item $right(n) \in V$, the right child of $n$ or $\emptyset$ if $n$ is a leaf. If $right(n) \neq \emptyset$, then $(n, right(n)) \in E$.
\end{enumerate}
We use $L \subseteq V$ to denote the set of leaves. % subseteq because tree could have single node

\subsection{Tiling}
\label{sec:Tiling}
% Treebeard vectorizes tree walks by grouping nodes of a decision tree into \textbf{\emph{tiles}}. The nodes in a tile are evaluated concurrently using vector instructions. Once the nodes of the current tile are evaluated, a look up table is used to compute which child of the current tile to move to next.

While a decision tree is naturally represented by a binary tree, 
it is not the best representation for tree traversal as it (i) requires many memory accesses, 
(ii) has poor branching structure and (iii) cannot make use of vector instructions. 
This section proposes a tiling optimization where we group multiple nodes in a decision tree into a single tile, effectively transforming a binary tree into an $n$-ary tiled tree. This not only allows the compiler to generate vectorized code (see Section~\ref{sec:Vectorization} ) to traverse trees but also enables spatial locality improvements by grouping nodes that are likely to be accessed together. We demonstrate this with two tiling heuristics later in the section.

\CommentOut{
Treebeard groups nodes of the decision tree into \textbf{\emph{tiles}}. Tiling provides two benefits. 
\begin{enumerate}
  \item It allows the compiler to generate vector code to traverse trees. Section \ref{sec:Vectorization} describes how Treebeard does this.
  \item It enables spatial locality improvements by grouping together nodes that are likely to be accessed together. 
\end{enumerate}
Once nodes are grouped into tiles, an $n$-ary tree whose nodes are tiles is constructed. Treebeard then generates 
optimized code to walk this tree. The listing below shows at a high level how a tiled tree is walked (This is not 
true IR, but presented for clarity). 
}

Once trees are tiled Treebeard generates tree walks with the code structure shown below.
\begin{lstlisting}[style=c++]
  ResultType Prediction_Function(...) {
    // ...
    Tile t = getRootTile(tree)
    while (!isLeaf(tree, t)) do {
      // Evaluate conditions of all nodes in the tile
      predicates = evaluateTilePredicates(t, rows[i])

      // Move to the correct child of the current tile
      t = getChildTile(tree, t, predicates) 
    }
    treePrediction = getLeafValue(t)
    // ...
  }  
\end{lstlisting}
The code is just an abstract representation of a tiled tree walk that enables efficient lowering of specific steps in subsequent stages. 
\op{evaluateTilePredicates} (speculatively) computes the predicates on all nodes in a tile (line 6). Then \op{getChildTile} (line 9), uses the computed predicate values to determine which child of the current tile to move to. We defer a description of how these operators are lowered to a later section but focus on tiling algorithms in this section.

\CommentOut{
To compute the prediction of the tree, the predicates of all nodes in the tile are computed simultaneously (line 6). 
Then, the computed predicate values are used to determine which child of the current tile to move to (in the 
$n$-ary tree). This section presents the details of tiling and how Treebeard's general tiling infrastructure 
can be used to develop tiling algorithms with different objectives (sections \ref{sec:UnifTiling}) and \ref{sec:ProbTiling}).
The details of how Treebeard lowers predicate evaluation and moving to the correct child to use vector instructions 
are described in section \ref{sec:Vectorization}.

\subsection{Tiles and Tree Tiling}
\label{sec:ValidTiling}
}
\subsection{Conditions for Tiling}
\label{sec:ValidTiling}
While any arbitrary partitioning of the nodes of a tree could be considered for tiling we impose a few intuitive constraints to only allow nodes that are likely to be accessed together to be grouped into a tile. 
\TODO{kr : replace $n_t$ with sz?}
Specifically given $tree = (V, E, r)$ and a tile size $n_t$ we impose the following constraints on the generated tiles $\{ T_1, T_2, ... ,T_m \}$ .
\begin{description}
    \item[Partitioning] $T_1 \cup T_2 ... \cup T_m = V$ and $T_i \cap T_j = \emptyset$ for all $i\neq j$
    \item[Connectedness] If $u, v \in T_i$, there is a (undirected) path connecting $u$ and $v$ fully contained in $T_i$.
   \item [Leaf seperation] $\forall l \in L$ : $l \in T_i \rightarrow v \notin T_i \;\; \forall v \in V \backslash \{l\}$
  \item [Maximal tiling] if there are tiles such that. $|T_i| < n_t$ then there is no $v \in V\backslash \{ T_i \cup L \}$ such that $(u, v) \in E$ for some $u \in T_i$. 
\end{description}
The \textbf{partitioning} and \textbf{maximal tiling} constraints together ensure that we group nodes into as few tiles as possible. {\textbf{Connectedness}} ensures that each tile is a sub-tree, a natural grouping of nodes that are accessed together. {\textbf{Leaf separation}} is critical to ensure that leaf checks can be efficiently performed.
\TODO{kr: improve the above}
\CommentOut{
\subsection{Tiled Trees}
A tiling transformation communicates the tiling to the Treebeard infrastructure by assigning a tile ID to each node in the decision tree. Using these tile IDs, Treebeard checks the validity of the tiling and then contructs a tree whose nodes are tiles. We call this tree the \textbf{\emph{tree of tiles}}. \TODO{We need a better name for this}
Figure \ref{Fig:ValidTilingTileSize3} shows a valid tiling with tile size 3 and the tree of tiles constructed by Treebeard. Three nodes are grouped into each of the tiles $t_1$ and $t_2$ as shown. Each tile is collapsed into a single node in the tree of tiles. However, each leaf in the original tree becomes a leaf in the tree of tiles.

@@

Treebeard maintains the following invariants.
\begin{enumerate}
  \item All tiles in a tree are the same size $n_t$. If the tiling produces any smaller tiles, these are padded by inserting dummy nodes to make them the required size.
  \item Nodes within tiles are always ordered in level order and left to right within a level. The numbering of the nodes in the above diagram shows this node order.
  \item Children of a node are numbered from left to right (regardless of level). For example, $l_1$ is the first child of $t_1$, $l_2$ is the second and so on.
\end{enumerate}
   }

### 2022-04-17 docs/MICRO 2022/uniftiling.tex
unif tililng

\subsubsection{More Notation}
In order to formulate the probability based tiling algorithm as an optimization problem, we define the following.
\begin{enumerate}
    \item For every leaf $l \in L$, we define $p_l$ as the probability that the leaf $l$ is reached.
    \item For each node $n \in V$, we define the absolute probability $p_v$ as
    \begin{equation}
        p_v = \begin{cases}
        p_l &\text{if $l \in L$}\\
        p_{left(v)} + p_{right(v)} &\text{otherwise}
        \end{cases}
    \end{equation}
    \item For any tree $T$, $\mathcal{C}(T)$ represents the set of all valid tilings of $T$.
    \item For every $v \in V$, we define $S_v$ as the subtree rooted at $v$.
    \item For every $v \in V$, we define $L_v$ as the set of leaves of $S_v$.
    \item For a every tile $T_i$, we define $root(T_i)$ as the node $v \in T_i$ such that $v$ has no incoming edges from any other node $u \in T_i$.
    \item For a tile $T_i$, $out(T_i) \subseteq E$ is the set of edges $(u, v)$ such that $u \in T_i$ and $v \notin T_i$.
\end{enumerate}

\subsection{Uniform Tiling}
\label{sec:UnifTiling}
% The justification for uniform tiling -- if we assume all paths are equally likely,
% we want to minimize the depth of speculation
The first tree tiling algorithm we implement is \textbf{\emph{uniform tiling}}. The core idea of uniform tiling is to construct tiles,
with a tile size $n_t$, so that we evaluate 
the least possible number of nodes speculatively. Intuitively, this translates to minimizing the height of the constructed tile. We minimize
the height of the tile by performing a level order traversal starting at the root node of the tile. 
\TODO{AP Should we say this is optimal when probabilities are equal? Problem is we haven't said what optimal is.}
This traversal collects at most $n_t$ nodes
and does not include any leaves in the tile so that the constraints listed in section \ref{sec:ValidTiling} are satisfied. Algorithm 
\ref{Alg:UnifTilingAlgo} lists the algorithm for uniform tiling. Tiling starts at the root and constructs a tile $Tile$ by performing
a level order traversal. Once the current tile is constructed, the tiling procedure is recursively performed on all nodes that are 
destinations for edges going out of the constructed tile. The set of tiles constructed by algorithm \ref{Alg:UnifTilingAlgo} constitutes 
a uniform tiling of the input tree.
% Algorithm for uniform tiling
\begin{algorithm}
  \caption{Uniform tree tiling}
  \label{Alg:UnifTilingAlgo}
  \begin{algorithmic}
      \Procedure{TileTree}{$T = (V, E, r)$, $n_t$} 
          \If {$r \in L$}
              \State \textbf{return} $\{ r \}$
          \EndIf
          \State \textcolor{codegreen}{\textit{//Level order traversal to collect $n_t$ or fewer nodes. }}
          \State \textcolor{codegreen}{\textit{//Leaves are not included in the constructed tile. }}
          \State $Tile \leftarrow LevelOrderTraveral(r, n_t)$
          \State $Tiles =  \{ Tile \}$
          \For{$(u,v) \in Out(Tile)$}
              \State $Tiles \leftarrow Tiles \cup TileTree(S_v, n_t)$
          \EndFor
          \State \textbf{return} $Tiles$
      \EndProcedure
  \end{algorithmic}
\end{algorithm}

>>>>

\subsection{Basic Tiling Algorithm}
\label{sec:UnifTiling}
% The justification for uniform tiling -- if we assume all paths are equally likely,
% we want to minimize the depth of speculation
Algorithm~\ref{Alg:UnifTilingAlgo} shows a simple tiling algorithm that satisfies all the above constraints. 
Tiling starts at the root and constructs a tile $Tile$ by performing
a level order traversal. The call \op{LevelOrderTraversal(r, $n_t$)} picks the next $n_t$ nodes according to the standard level order tree traversal algorithm (not described here) applied only to internal nodes of the tree. Once the current tile is constructed, the tiling procedure is recursively performed on all nodes that are 
destinations for edges going out of the constructed tile.
It is easy to see that the set of tiles constructed by algorithm \ref{Alg:UnifTilingAlgo} satisfies the constraints above.
\TODO{Pass a modified tree without leafs to the level order traversal call. Explicitly add leaves as seperate tiles}
% Algorithm for uniform tiling
\begin{algorithm}
  \caption{Uniform tree tiling}
  \label{Alg:UnifTilingAlgo}
  \begin{algorithmic}
      \Procedure{TileTree}{$tree = (V, E, r)$, $n_t$} 
          \If {$r \in L$}
              \State \textbf{return} $\{ r \}$
          \EndIf
          \State \textcolor{codegreen}{\textit{//Level order traversal to collect $n_t$ or fewer nodes. }}
          \State \textcolor{codegreen}{\textit{//Leaves are not included in the constructed tile. }}
          \State $Tile \leftarrow LevelOrderTraveral(r, n_t)$
          \State $Tiles =  \{ Tile \}$
          \For{$(u,v) \in Out(Tile)$}
              \State $Tiles \leftarrow Tiles \cup TileTree(S_v, n_t)$
          \EndFor
          \State \textbf{return} $Tiles$
      \EndProcedure
  \end{algorithmic}
\end{algorithm}
